---
title: "test"
author: "SP"
date: "2/17/2023"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r setup, echo = FALSE, include=FALSE}
library(tidyverse)
library(tidycensus)
library(tigris)
library(sf)
library(ggspatial)
library(gridExtra)
library(RColorBrewer)
library(reshape2)
library(tidytransit)
library(naniar)

#counties stored as list
counties <- c("Oklahoma", "Canadian", "Cleveland", "Grady", "Lincoln", "Logan", "McClain")

#desired ACS variables stored as list
hh_vars <- c(
            # total pop
            tot_pop = "B01003_001",
      
            # vehicle availability
            no_veh = "B08201_002",
            
            #general household statistics
            total_hhs = "B08201_001",
            hh_1person = "B08201_007",
            hh_2person = "B08201_013",
            hh_3person = "B08201_019",
            hh_4person_plus = "B08201_025",
            
            #household structure
            hh_u18_married_couple = "B09002_002",
            hh_u18_singleparent_male = "B09002_009",
            hh_u18_singleparent_female = "B09002_015",
            hh_65plus = "B11007_002",
            
            #disability
            tot_disabled = "B18101_001",
            
            #income statistics
            inc_lt_10k = "B19001_002",
            inc_btw_10k_15k = "B19001_003",
            inc_btw_15k_20k = "B19001_004",
            inc_btw_20k_25k = "B19001_005",
            inc_btw_25k_30k = "B19001_006",
            inc_btw_30k_35k = "B19001_007",
            inc_btw_35k_40k = "B19001_008",
            inc_btw_40k_45k = "B19001_009",
            inc_btw_45k_50k = "B19001_010",
            inc_btw_50k_60k = "B19001_011",
            inc_btw_60k_75k = "B19001_012",
            inc_btw_75k_100k = "B19001_013",
            inc_btw_100k_125k = "B19001_014",
            inc_btw_125k_150k = "B19001_015",
            inc_btw_150k_200k = "B19001_016",
            inc_gt_200k = "B19001_017",
            hh_povlevel = "B17019_002")
# Define color palettes
blue_pal <- brewer.pal(5, "Blues")
red_pal <- brewer.pal(5, "Reds")
purp_pal <- brewer.pal(5, "Purples")
green_pal <- brewer.pal(5,"Greens")
orange_pal <- brewer.pal(5,"Oranges")
```

```{r get_data, include=FALSE}
# get ACS data from census API
census_data <- get_acs(geography = "tract",
                              state = "OK",
                              county = counties,
                              variables = hh_vars,
                              output = "wide", 
                              year = 2019,
                              geometry = TRUE) 
# get land area from tigris
okc_areas <- tracts(state = "OK",
                    county = counties,
                    year = 2019) %>%
              select(GEOID, ALAND) %>%
              st_drop_geometry()

# get employment data from LEHD (all OK)
lehd_data <- read_csv("https://lehd.ces.census.gov/data/lodes/LODES7/ok/wac/ok_wac_S000_JT00_2019.csv.gz", show_col_types = FALSE) %>%
  rename(total_emp = C000) %>%
  mutate(basic_emp = CNS01+CNS02+CNS03+CNS04+CNS05+CNS06+CNS08+CNS09) %>%
  rename(retail_emp = CNS07) %>%
  mutate(service_emp = total_emp - basic_emp - retail_emp) %>%
  select(w_geocode, total_emp, basic_emp, retail_emp, service_emp) %>%
  mutate(w_geocode = as.character(w_geocode)) %>%
  mutate(GEOID = substr(w_geocode,1,11)) %>% 
  select(-w_geocode) %>%
  group_by(GEOID) %>% 
  summarise(across(everything(), ~sum(.)))

# join all data frames
ok_msa <- census_data %>% 
  left_join(lehd_data) %>%
  left_join(okc_areas) %>%
  select(-ends_with("M")) %>%
  rename(land_area_sqmeters = ALAND) %>%
  st_transform(crs = 4326)

#compute density variables
ok_msa <- ok_msa %>%
  mutate(pop_density = tot_popE/land_area_sqmeters) %>%
  mutate(emp_density = total_emp/land_area_sqmeters) %>%
  mutate(activity_density = (tot_popE + total_emp)/land_area_sqmeters)
```
# Data Overview
The following data was aggregated and computed from the 2021 ACS 5-year estimate, and Longitudinal Employer-Household Dynamica dataset. 
An overview of the resulting dataset is provided below:
## Data Structure:
The sample table below shows the first 5 rows of our compiled dataset. Each row represents a census tract in the study area along with its corresponding variables and geometry. When necessary, this dataset can easily be grouped by county, pivoted by category, or mapped geographically. 
```{r summary table, echo=FALSE}
tibble(ok_msa[1:5,])
```
## Data Definitions:\
**GEOID** - Numerical ID for census tract                	
**NAME** - Census tract, county, and state                    	
**tot_popE** - Total number of people    	
**no_vehE** - Households with no vehicle present                	
**total_hhsE** - Total number of households               	
**hh_1personE** - Number of households with 1 person           	 
**hh_2personE** - Number of households with 2 persons                 	
**hh_3personE** - Number of households with 3 persons                	
**hh_4person_plusE** - Number of households with 4 or more persons           	
**hh_u18_married_coupleE** - Number of households with a person under 18 and a married couple head-of-household  	
**hh_u18_singleparent_maleE** - Number of households with a person under 18 and a single male head-of-household  	
**hh_u18_singleparent_femaleE** - Number of households with a person under 18 and a single female head-of-household  	
**hh_65plusE** - Number of households with a person age 65 or older             	
**tot_disabledE** - Number of people with a disability           	
**inc_lt_10kE** - Number of households with income less than $10,000            	
**inc_btw_10k_15kE** - Households with income between $10,000 and $15,000       	
**inc_btw_15k_20kE** - Households with income between $15,000 and $20,000       	
**inc_btw_20k_25kE** - Households with income between $20,000 and $25,000    	 
**inc_btw_25k_30kE** - Households with income between $25,000 and $30,000      	
**inc_btw_30k_35kE** - Households with income between $30,000 and $35,000      	
**inc_btw_35k_40kE** - Households with income between $35,000 and $40,000       	
**inc_btw_40k_45kE** - Households with income between $40,000 and $45,000       	
**inc_btw_45k_50kE** - Households with income between $45,000 and $50,000       	
**inc_btw_50k_60kE** - Households with income between $50,000 and $60,000      	 
**inc_btw_60k_75kE** - Households with income between $60,000 and $75,000         	
**inc_btw_75k_100kE** - Households with income between $75,000 and $100,000  
**inc_btw_100k_125kE** - Households with income between $100,000 and $125,000      	
**inc_btw_125k_150kE** - Households with income between $125,000 and $150,000  	
**inc_btw_150k_200kE** - Households with income between $150,000 and 200,000	
**inc_gt_200kE** - Households with income greater than $200,000         	 
**hh_povlevelE** - Number of households below the poverty level            	
**total_emp** - Total number of people employed            
**basic_emp** - Total number of people employed in the following sectors: 
    Agriculture, Forestry, Fishing, and Hunting (CNS01)
    Mining and extraction (CNS02)
    Utilities (CNS03)
    Construction (CNS04)
    Manufacturing (CNS05)
    Wholesale trade (CNS06)
    Transportation and warehousing (CNS06)  
**retail_emp** - Total number of people employed in retail             	
**service_emp** - Total number of people employed in remaining sectors            	
**land_area_sqmeters** - Land area in square meters   	 
**geometry** - Geographic coordinates of census tract outlines                	
**pop_density** - People per square meter (calculated)           
**emp_density** - Employees per square meter (calculated)                	
**activity_density** - People and employees per square meter (calculated)
# Geography Overview
The Oklahoma City metropolitan statistical area (MSA) is composed of seven counties centrally located in Oklahoma: Canadian County, Cleveland County, Grady County, Lincoln County, Logan County, McClain County, and Oklahoma County. Across these counties there are a total of **363 census tracts** (based on the 2020 redistricting). The MSA has a total land area of approximately **1,427,523 square kilometers**.Together, these counties are home to 1,412,900 people, according to the 2021 ACS 5-Year Estimates. The Oklahoma City MSA makes up nearly 36% of the state’s total population of 3,948,100. 
The Oklahoma City MSA is predominantly white, with 63% of residents identifying as white alone, followed by 13.9% Hispanic or Latino, 10% Black, 3.1% American Indian and Alaska Native, 3.1% Asian, 0.1% Native Hawaiian and Pacific Islander. 37% of residents in the study are are ages 35-64, 24.6% are under the age of 18, 24.4% are ages 18-34, and 14% are ages 65 and older. The median household income in 2021 was $63,351 – higher than the statewide average and lower than the national average. Just over 14% of residents live below the poverty line – lower than the statewide average and higher than the national average. 
## Transit 
EMBARK, the area’s transit authority, operates all public transit in greater Oklahoma City, which includes fixed-route bus service, the OKC Streetcar, paratransit service, river ferry transit, and a bikeshare network. Beyond transit, car-ownership and use is prevalent in the Oklahoma City MSA with only 2.6% of households reporting that they do not have access to a vehicle, and a median number of vehicles per household of 2.3. According to the 2021 ACS 5-Year Estimates, cars, trucks, or vans are the most common means of transportation to work for workers 16 and over (used by 89.7% of respondents), while commuting by public transportation is far less common (used by 0.5% of respondents). Based on the same ACS data, the average travel time to work is 23 minutes. 
## Employment
Major employment sectors in the Oklahoma City metro area include government, higher education, aerospace, healthcare, and retail. According to the Greater Oklahoma City Chamber of Commerce, major employers include the State of Oklahoma, Tinker Air Force Base, the University of Oklahoma, Integris Health, and Amazon. In 2021, the metropolitan area added 10,825 jobs (1.7% increase), and further job growth was expected in 2022 as stated in the 2022 Greater Oklahoma City Economic Outlook.  
```{r geog, fig.height = 6.5, fig.width = 6.5, fig.align = "center", fig.cap="\\label{fig:figs}Map showing the counties and census tracts in the Oklahoma City MSA", echo=FALSE}
county_geom <- ok_msa %>%
    separate(NAME, into = c("tract-num", "county", "state"), sep = ", ") %>%
    group_by(county) %>%
    summarize(geometry = st_union(geometry))
ggplot() +
  annotation_map_tile(zoomin = 0, progress = "none", type = "cartolight") +
  geom_sf(data = ok_msa, aes(), fill = NA, color="darkgrey", size=0.1) +
  geom_sf(data = county_geom, aes(), fill = NA, color="black", size=0.5) +
  geom_sf_label(data = county_geom, aes(label = county), alpha = 0, label.size = NA, size =4, fontface = "bold") +
  annotation_scale(location = 'bl', text_cex = 1) +
  theme_void() 
```
  
## Densities  
```{r pop_density_map,  message=FALSE, fig.height = 3.5, fig.align = "left", fig.cap="\\label{fig:figs}Map showing the population density in the Oklahoma City MSA", echo=FALSE}
pop_density_map <- ggplot(ok_msa) +
  annotation_map_tile(zoomin = 0, progress = "none", type = "cartolight") +
  geom_sf(aes(fill = pop_density), color=NA, size=0.1) +
  geom_sf(data = county_geom, aes(), fill = NA, color="black", size=0.1) +
  scale_fill_gradientn(colors = blue_pal, name = "People / square meter", labels = paste(seq(0,4)/1000)) +
  scale_fill_gradientn(colors = blue_pal, trans = "log", name = "People / square meter", labels = paste(seq(0,3)*.001)) +
  theme_void()
pop_density_map
```
```{r emp_density_map, fig.height = 3.5, fig.align = "left", fig.cap="\\label{fig:figs}Map showing the employment density in the Oklahoma City MSA", echo=FALSE}
emp_density_map <- ggplot(ok_msa) +
  annotation_map_tile(zoomin = 0, progress = "none", type = "cartolight") +
  geom_sf(aes(fill = emp_density), color=NA, size=0.1) +
  geom_sf(data = county_geom, aes(), fill = NA, color="black", size=0.1) +
  scale_fill_gradientn(colors = red_pal, name = "Jobs / square meter", labels = paste(seq(0,4)*.001)) + 
  scale_fill_gradientn(colors = red_pal, trans = "log", name = "Jobs / square meter", labels = paste(seq(0,4)*.001)) + 
  theme_void()
emp_density_map
```
  
## Employment & Income  
```{r employemnt chart, fig.height = 8, fig.width = 12, fig.align = "center", fig.cap="\\label{fig:figs}Employment by sector by county", echo=FALSE}
county_employment <- ok_msa %>%
    separate(NAME, into = c("tract-num", "county", "state"), sep = ", ") %>%
    group_by(county) %>%
    summarise(
      basic_emp = sum(basic_emp),
      service_emp = sum(service_emp),
      retail_emp = sum(retail_emp)) %>%
    st_drop_geometry()
county_employment_stacked <- melt(county_employment, id = "county")
ggplot(county_employment_stacked, aes(x = county, y = value, fill = variable)) + 
  geom_bar(stat = "identity") +
  scale_fill_manual(values = c("steelblue","thistle", "gold"), label = c("Basic", "Service", "Retail")) +
  scale_y_continuous(labels = scales::comma) +
  labs(y = "Number of Employees", x = " ", fill = "Employment Sector") + 
  theme(legend.position = "top")
```
  
```{r income_distribution, fig.height = 6, fig.width = 12, fig.align = "center", fig.cap="\\label{fig:figs}Distribution of households by income bracket", echo=FALSE}
msa_income <- ok_msa %>%
  separate(NAME, into = c("tract-num", "county", "state"), sep = ", ") %>%
    group_by(state) %>%
    summarise(
      inc_lt_10kE = sum(inc_lt_10kE),
      inc_btw_10k_15kE = sum(inc_btw_10k_15kE),
      inc_btw_15k_20kE = sum(inc_btw_15k_20kE),
      inc_btw_20k_25kE = sum(inc_btw_20k_25kE),
      inc_btw_25k_30kE = sum(inc_btw_25k_30kE),
      inc_btw_30k_35kE = sum(inc_btw_30k_35kE),
      inc_btw_35k_40kE = sum(inc_btw_35k_40kE),
      inc_btw_40k_45kE = sum(inc_btw_40k_45kE),
      inc_btw_45k_50kE = sum(inc_btw_45k_50kE),
      inc_btw_50k_60kE = sum(inc_btw_50k_60kE),
      inc_btw_60k_75kE = sum(inc_btw_60k_75kE),
      inc_btw_75k_100kE = sum(inc_btw_75k_100kE),
      inc_btw_100k_125kE = sum(inc_btw_100k_125kE),
      inc_btw_125k_150kE = sum(inc_btw_125k_150kE),
      inc_btw_150k_200kE = sum(inc_btw_150k_200kE),
      inc_gt_200kE = sum(inc_gt_200kE)) %>%
  pivot_longer(
    cols = starts_with("inc"),
    values_to = "count"
  ) %>% 
  rename(inc_level = name) %>%
  select(inc_level, count)
msa_income$inc_level <- factor(msa_income$inc_level, levels = msa_income$inc_level)
ggplot(msa_income) +
  geom_col(aes(x = inc_level, y = count), fill = "black") +
  scale_y_continuous(labels = scales::comma) +
  labs(y = "Number of Households", x = "Income Bracket") +
  scale_x_discrete(labels=c('<10k', '10-15k', '15-20k', "20-25k", "25-30k", "30-35k", "35-40k", "40-45k", "45-50k", "50-60k", "60-75k", "75-100k", "100-125k", "125-150k", "150-200k", ">200k"))
```

```{r poverty_map, fig.height = 4, fig.align = "left", fig.cap="\\label{fig:figs}Map showing the spatial distribution of poverty in the Oklahoma City MSA", echo=FALSE}
ggplot(ok_msa) +
  annotation_map_tile(zoomin = 0, progress = "none", type = "cartolight") +
  geom_sf(aes(fill = (hh_povlevelE/total_hhsE)), color=NA, size=0.1) +
  scale_fill_gradientn(colors = green_pal, name = "% of households below federal poverty line", labels = paste(seq(0,5),"%")) + 
  theme_void()
```
  
## Vehicle Ownership  
```{r vehicle_ownership_map, fig.height = 4, fig.align = "left", fig.cap="\\label{fig:figs}Map showing the spatial distribution of vehicle access in the Oklahoma City MSA", echo=FALSE}
pct_vehicle_ownership <- ok_msa %>%
  mutate(pct_veh = (1-(no_vehE/total_hhsE))) %>%
  select(GEOID, NAME, no_vehE, total_hhsE, pct_veh, geometry)
veh_distribution_map <- ggplot(pct_vehicle_ownership) +
  annotation_map_tile(zoomin = 0, progress = "none", type = "cartolight") +
  geom_sf(aes(fill = pct_veh), color=NA, size=0.1) +
  scale_fill_gradientn(colors = orange_pal, name = "% of households with 1+ vehicle", labels = paste(seq(4,10, by=2)*10,"%")) +
  theme_void()
veh_distribution_map
```

```{r vehicle_ownership_hist, fig.height = 4.5, fig.width = 10, fig.align = "left", fig.cap="\\label{fig:figs}Distribution of cencus tracts by percent of households with access to a vehicle", echo=FALSE}
veh_distribution_chart <- ggplot(pct_vehicle_ownership) +
  geom_histogram(aes(x = pct_veh), fill = "orangered3", bins = 50) + 
  labs(x = "% of households with 1+ vehicle", y = "Number of Census Tracts") +
  scale_x_continuous(labels = scales::percent)
veh_distribution_chart
```

```{r write to file, echo=FALSE}
#zone_boundaries <- ok_msa %>%
  #select(GEOID, geometry)
#st_write(zone_boundaries, "zonesboundaries.geojson", append = FALSE)
```
# Travel Time Matrix: Roads
To create the network in TransCAD, we imported the Open Street Maps layer and selected for primary, secondary, tertiary, and trunk roads. When creating the centroids and centroid connectors, we allowed for the connectors to extend outside of zone boundaries, at up to 30 miles, and up to 10 connectors per centroid. This resulted in a matrix with only a few gaps, which we resolved by turning surrounding roads two-way. (The major roads in the area were separated highways, so we trust that two-way driving is actually feasible in these zones.) 
```{r read in TransCAD matrix, echo=FALSE}
#read in drive time matrix
ok_matrix_drive_path <- "../Travel-Forecasting/TTmatrix_skim10.csv"
ok_matrix_drive <- read.csv(ok_matrix_drive_path)

#read in drive time centroids
ok_nodes_centroids_path <- "../Travel-Forecasting/nodes_okc_roads_prayerhands_withcentroids.csv"
#convert lat and long units
ok_nodes_centroids <- read.csv(ok_nodes_centroids_path) %>%
  drop_na(centroid) %>%
  mutate(long = Longitude / 1000000) %>%
  mutate(lat = Latitude / 1000000)
#create drive time centroid geometry
ok_centroids_geom <- st_as_sf(ok_nodes_centroids, 
                                     coords=c("long", "lat"), 
                                     crs = 4326,
                                     agr = "constant") %>%
                      select(ID, geometry)
```

```{r mapping TransCAD file, echo=FALSE}
#ggplot(ok_msa) +
  #geom_sf(aes(), fill="white") +
  #geom_sf(data = okc_centroids_geom, fill = "black")
```

```{r join TransCAD files, echo=FALSE}
#ok_msa_w_centroids <- ok_msa
#ok_msa_w_centroids$centroid_id <- apply(st_contains(ok_msa, ok_centroids_geom, sparse = FALSE), MARGIN = 2, 
               #function(col) { 
                #  ok_centroids_geom[which(col), ]$ID
               ##st_write(ok_msa_w_centroids, "ok_msa_w_centroids.csv", append = FALSE)

#pivot drive time matrix
drive_TTmatrix_long <- ok_matrix_drive %>%
   pivot_longer(
     cols = starts_with("X"),
     values_to = "travel_time_mins") %>%
  #rename(origin = ï..Origin, destination = name) %>%
  rename(origin = Origin, destination = name) %>%
  mutate(destination = as.numeric(substr(destination, 2, 7))) %>%
  filter(travel_time_mins != "--") %>%
  mutate(travel_time_mins = as.numeric(travel_time_mins))
```
## Isochrones
After generating a complete matrix, we read the travel time matrix and the nodes layer into R. We converted the centroids’ coordinates to a geometry, matched centroid IDs with the appropriate census tracts, and pivoted the travel time matrix so origins and destinations would both be in columns. This allowed us to create an isochrone illustrating the travel time from one zone in the study area to all others. 

```{r main data frame with GEOIDs and drive+transit centroids, echo=FALSE}
#read in updated data frame with GEOIDs, drive centroids, and transit centroids
GEOID_centroids_path <- "../Travel-Forecasting/ok_msa_w_centroids_updated_2-18.csv"
GEOID_centroids <- read.csv(GEOID_centroids_path) %>%
  rename(drive_centroid_id = centroid_id, 
         GEOID = ï..GEOID) %>%
         #GEOID = GEOID) %>%
  mutate(GEOID = as.character(GEOID))
#join ok msa df with drive+transit centroids
ok_msa_w_centroids <- ok_msa %>%
  left_join(GEOID_centroids)
```

```{r drive time isochrone generator, echo=FALSE}
drive_isochrone_generator <- function(iso_origin){
  #create temp matrix to isolate one origin
  temp_matrix <- drive_TTmatrix_long %>%
  filter(origin == iso_origin)
  #join ok msa w centroids to temp matrix to give geometry
  iso_drivetime_from_origin <- ok_msa_w_centroids %>%
  select(GEOID, geometry, drive_centroid_id) %>%
  left_join(temp_matrix, by = c("drive_centroid_id" = "destination"))
  #assign NA = 0 for travel time
iso_drivetime_from_origin[is.na(iso_drivetime_from_origin)] <- 0
#isolate centroid geometry to identify one point
iso_origin_geom <- ok_centroids_geom %>%
  filter(ID == iso_origin)
#plot zone geometry with drive time mins as fill
ggplot(iso_drivetime_from_origin) + 
  geom_sf(aes(fill = travel_time_mins)) + 
  scale_fill_gradient2(low = "white", mid = "thistle", high = "orchid4",
                       midpoint = 50, #median(iso_drivetime_from_origin$travel_time_mins),
                       name = paste("Drive Time\nto all\nzones from", iso_origin, "\n(minutes)")) +
  geom_sf(data = iso_origin_geom, shape = 19, size = 1.5)
}
```

```{r drive time isochrones, echo=FALSE}
drive_isochrone_generator(848969)

drive_isochrone_generator(848732)

drive_isochrone_generator(848750)

transit_isochrone_generator(41038)
```

## Accessibility Score

This metric allows us to compare the relative accessibility of tracts more easily. We calculated an "accessibility score" by counting the number of zones each origin could reach within 30 minutes. We note that this score is affected by the lack of residential roads in our network, which would likely provide better accessibility for some of the peripheral zones. 

```{r accessibility score, echo=FALSE}
drive_accessibility_matrix <- drive_TTmatrix_long %>%
  filter(travel_time_mins<30) %>%
  group_by(origin) %>%
  tally() %>%
  rename(score = n)

drive_accessibility_score_geom <- ok_msa_w_centroids %>%
  select(GEOID, NAME, geometry, drive_centroid_id) %>%
  left_join(drive_accessibility_matrix, by = c("drive_centroid_id" = "origin"))
 
ggplot(drive_accessibility_score_geom) + 
  geom_sf(aes(fill=score))
```

# Travel Time Matrix: Transit

## Assumptions:

In our model and in this house we assume: everyone walks to transit, vehicle speed is 23 mph, walking speed is 2.5 mph, 15-minute wait time, fare is 2 dollars and it is free to transfer.

```{r read in transit skim matrices and route shapes, echo=FALSE}
transitskim_fare_path <- "../Travel-Forecasting/transit_skim1_fare.csv"
transitskim_fare <- read.csv(transitskim_fare_path)
transitskim_ivtt_path <- "../Travel-Forecasting/transit_skim1_ivtt.csv"
transitskim_ivtt <- read.csv(transitskim_ivtt_path)
transitskim_init_wt_path <- "../Travel-Forecasting/transit_skim1_init_waittime.csv"
transitskim_init_wt <- read.csv(transitskim_init_wt_path)
transitskim_transfer_wt_path <- "../Travel-Forecasting/transit_skim1_transfer_waittime.csv"
transitskim_tranfer_wt <- read.csv(transitskim_transfer_wt_path)
transitskim_access_wt_path <- "../Travel-Forecasting/transit_skim1_access_walktime.csv"
transitskim_access_wt <- read.csv(transitskim_access_wt_path)
transitskim_egress_wt_path <- "../Travel-Forecasting/transit_skim1_egress_walktime.csv"
transitskim_egress_wt <- read.csv(transitskim_egress_wt_path)
transitskim_totaltime_path <- "../Travel-Forecasting/transit_skim1_total_time.csv"
transitskim_totaltime <- read.csv(transitskim_totaltime_path)
transitskim_no_transfers_path <- "../Travel-Forecasting/transit_skim1_no_transfers.csv"
transitskim_no_transfers <- read.csv(transitskim_no_transfers_path)

okc_gtfs <- read_gtfs("https://embarkok.com/data/gtfs/google_transit.zip")
route_shapes <- shapes_as_sf(okc_gtfs$shapes)
```

```{r reading transit centroids, include = FALSE}
#read in transit centroids
ok_matrix_transit_path <- "../Travel-Forecasting/Nodes_AND_area_centroids.csv"
#convert lat and long units
ok_matrix_transit_centroids <- read.csv(ok_matrix_transit_path) %>% 
  drop_na(centroid) %>%
  mutate(long = Longitude / 1000000) %>%
  mutate(lat = Latitude / 1000000)
#create transit centroid geometry
ok_transit_centroids_geom <- st_as_sf(ok_matrix_transit_centroids, 
                                     coords=c("long", "lat"), 
                                     crs = 4326,
                                     agr = "constant") %>%
                      select(GEOID, ID) %>%
                      mutate(GEOID = as.character(GEOID)) 

#ok_transit_centroids <- ok_transit_centroids_geom %>%
                      #st_drop_geometry()

#ok_msa_w_centroids <- left_join(ok_msa_w_centroids, ok_transit_centroids) %>%
  #rename(transit_centroid_id = ID)
```

```{r transit vizzz, include = FALSE}
#pivot transit skims
transit_totaltime_long <- transitskim_totaltime %>%
  pivot_longer(
    cols = starts_with("X"),
    values_to = "total_time_mins") %>%
  #rename(origin = ï..RCIndex, destination = name) 
  rename(origin = RCIndex, destination = name) %>%
  mutate(destination = as.numeric(substr(destination, 2, 6))) %>%
  replace_with_na(replace = list(x = "--")) %>%
  mutate(total_time_mins = as.numeric(total_time_mins))

transit_fare_long <- transitskim_fare %>%
  pivot_longer(
    cols = starts_with("X"),
    values_to = "fare") %>%
 #rename(origin = ï..RCIndex, destination = name) 
  rename(origin = RCIndex, destination = name) %>%
  mutate(destination = as.numeric(substr(destination, 2, 6))) %>%
  replace_with_na(replace = list(x = "--")) %>%
  mutate(fare = as.numeric(fare))

transit_ivtt_long <- transitskim_ivtt %>%
  pivot_longer(
    cols = starts_with("X"),
    values_to = "ivtt") %>%
  #rename(origin = ï..RCIndex, destination = name) 
  rename(origin = RCIndex, destination = name) %>%
  mutate(destination = as.numeric(substr(destination, 2, 6))) %>%
  replace_with_na(replace = list(x = "--")) %>%
  mutate(ivtt = as.numeric(ivtt))

transit_init_wt_long <- transitskim_init_wt %>%
  pivot_longer(
    cols = starts_with("X"),
    values_to = "init_wt") %>%
  #rename(origin = ï..RCIndex, destination = name) 
  rename(origin = RCIndex, destination = name) %>%
  mutate(destination = as.numeric(substr(destination, 2, 6))) %>%
  replace_with_na(replace = list(x = "--")) %>%
  mutate(init_wt = as.numeric(init_wt))

transit_transfer_wt_long <- transitskim_tranfer_wt %>%
  pivot_longer(
    cols = starts_with("X"),
    values_to = "transfer_wt") %>%
  #rename(origin = ï..RCIndex, destination = name) 
  rename(origin = RCIndex, destination = name) %>%
  mutate(destination = as.numeric(substr(destination, 2, 6))) %>%
  replace_with_na(replace = list(x = "--")) %>%
  mutate(transfer_wt = as.numeric(transfer_wt))

transit_access_wt_long <- transitskim_access_wt %>%
  pivot_longer(
    cols = starts_with("X"),
    values_to = "access_wt") %>%
  #rename(origin = ï..RCIndex, destination = name) 
  rename(origin = RCIndex, destination = name) %>%
  mutate(destination = as.numeric(substr(destination, 2, 6))) %>%
  replace_with_na(replace = list(x = "--")) %>%
  mutate(access_wt = as.numeric(access_wt))

transit_egress_wt_long <- transitskim_egress_wt %>%
  pivot_longer(
    cols = starts_with("X"),
    values_to = "egress_wt") %>%
  #rename(origin = ï..RCIndex, destination = name) 
  rename(origin = RCIndex, destination = name) %>%
  mutate(destination = as.numeric(substr(destination, 2, 6))) %>%
 replace_with_na(replace = list(x = "--")) %>%
  mutate(egress_wt = as.numeric(egress_wt))

transit_no_transfers_wt_long <- transitskim_no_transfers %>%
  pivot_longer(
    cols = starts_with("X"),
    values_to = "num_transfers") %>%
  #rename(origin = ï..RCIndex, destination = name) 
  rename(origin = RCIndex, destination = name) %>%
  mutate(destination = as.numeric(substr(destination, 2, 6))) %>%
  replace_with_na(replace = list(x = "--")) %>%
  mutate(num_transfers = as.numeric(num_transfers))

transit_TTmatrix_long <- full_join(transit_totaltime_long, transit_fare_long) %>% 
  full_join(transit_ivtt_long) %>% 
  full_join(transit_init_wt_long) %>% 
  full_join(transit_transfer_wt_long) %>% 
  full_join(transit_access_wt_long) %>% 
  full_join(transit_egress_wt_long) %>% 
  full_join(transit_no_transfers_wt_long) %>%
  mutate(ovtt = total_time_mins - ivtt) %>%
  mutate(perceived_time = ivtt + (1.5 * ovtt)) %>%
  select(origin, destination, total_time_mins, num_transfers, access_wt, init_wt, ivtt, transfer_wt, egress_wt, ovtt, perceived_time, fare)
```

## Plotting EMBARK Routes

This includes both bus routes and a streetcar route that we coded as rail.

```{r plotting EMBARK routes, echo=FALSE}
ggplot() +
  geom_sf(data = ok_msa,
          fill = "cornsilk",
          color = "grey") +
  geom_sf(data = route_shapes,
          aes(color = shape_id)) +
          theme_void()+
  theme(legend.position = "none")
  
```

## Transit Isochrones and Travel Time Break Down

Advantage of public transit data is break down the overall trip and travel time to produce a more granular analysis. The shape of the first charts are similar, demonstrating similar travel patterns in the downtown region. The third chart illustrates a much steeper increase in travel time. We ordered the stacked bar chart from shortest to longest amount of time rather than geographic distance. It is interesting to compare the three stacked bar charts. The first two follow a more exponential trend while the third follows more of a logarithmic trend.

```{r isochrone transit generator, include=FALSE}
transit_isochrone_generator <- function(iso_origin){
  temp_matrix <- transit_TTmatrix_long %>%
  filter(origin == iso_origin)
  
iso_transittime_from_origin <- ok_msa_w_centroids %>%
  select(GEOID, NAME, geometry, transit_centroid_id) %>%
  left_join(temp_matrix, by = c("transit_centroid_id" = "destination"))

iso_origin_geom <- ok_transit_centroids_geom %>%
  filter(ID == iso_origin)

ggplot(iso_transittime_from_origin) + 
  geom_sf(aes(fill = total_time_mins)) + 
  scale_fill_gradient2(low = "white", mid = "thistle", high = "orchid4",
                       midpoint = 50,
                       name = paste("Transit Time\nto all\nzones from", iso_origin, "\n(minutes)")) +
  geom_sf(data = iso_origin_geom, shape = 19, size = 1.5)
}
```

```{r transit stacked bar chart generator, include=FALSE}
transit_stackedbar_generator <- function(iso_origin){
  temp_matrix <- transit_TTmatrix_long %>%
  filter(origin == iso_origin) %>%
  select(destination, access_wt, init_wt, ivtt, transfer_wt, egress_wt)
temp_matrix_stacked <- melt(temp_matrix, id = "destination")
ggplot(temp_matrix_stacked, aes(x = reorder(destination, value), y = value, fill = variable)) + 
  geom_bar(stat = "identity") +
  scale_fill_manual(values = c("steelblue","thistle", "gold", "red", "green"), label = c("Walk time", "Wait time", "Travel time", "Transfer time", "Egress time")) +
  scale_y_continuous(labels = scales::comma) +
  labs(y = "Number of Minutes", x = " ", fill = "Public Transportation Activity", title = paste("Transit Time to all zones from", iso_origin, "(minutes)"))  + 
  theme(legend.position = "top", axis.text.x=element_blank(), axis.ticks.x=element_blank())
}
```

```{r plots, echo=FALSE, warning=FALSE}
transit_isochrone_generator(41028)
transit_stackedbar_generator(41028)
transit_isochrone_generator(41077)
transit_stackedbar_generator(41077)
transit_isochrone_generator(41097)
transit_stackedbar_generator(41097)
```

# Accessibility 

```{r adding GEOID to transit skim, echo=FALSE}
partial_drive_TTmatrix_long <- drive_TTmatrix_long %>%
  left_join(GEOID_centroids, by = c("origin" = "drive_centroid_id")) %>%
  rename(from_GEOID = GEOID) %>%
  select(from_GEOID, destination, travel_time_mins) %>%
  left_join(GEOID_centroids, by = c("destination" = "drive_centroid_id")) %>%
  rename(to_GEOID = GEOID, 
         drive_time = travel_time_mins) %>%
  select(from_GEOID, to_GEOID, drive_time)

partial_transit_TTmatrix_long <- transit_TTmatrix_long %>%
  left_join(GEOID_centroids, by = c("origin" = "transit_centroid_id")) %>%
  rename(from_GEOID = GEOID) %>%
  select(-origin) %>%
  left_join(GEOID_centroids, by = c("destination" = "transit_centroid_id")) %>%
  rename(to_GEOID = GEOID) %>%
  select(from_GEOID, to_GEOID, total_time_mins, num_transfers, access_wt, init_wt, ivtt, transfer_wt, egress_wt, ovtt, perceived_time, fare)

full_TTmatrix_long <- left_join(partial_drive_TTmatrix_long, partial_transit_TTmatrix_long, by = c("from_GEOID", "to_GEOID"))
```

```{r isochrone complete transportation generator, include=FALSE}
transpo_isochrone_generator <- function(iso_origin){
  temp_matrix <- full_TTmatrix_long %>%
  filter(from_GEOID == iso_origin)
  
iso_transpo_time_from_origin <- ok_msa_w_centroids %>%
  select(GEOID, NAME, geometry) %>%
  left_join(temp_matrix, by = c("GEOID" = "from_GEOID"))

iso_origin_geom <- ok_transit_centroids_geom %>%
  filter(ID == iso_origin)

ggplot(iso_transittime_from_origin) + 
  geom_sf(aes(fill = total_time_mins)) + 
  scale_fill_gradient2(low = "white", mid = "thistle", high = "orchid4",
                       midpoint = 50,
                       name = paste("Transit Time\nto all\nzones from", iso_origin, "\n(minutes)")) +
  geom_sf(data = iso_origin_geom, shape = 19, size = 1.5)
}
```


```{r creating centroid info}

complete_centroid <- ok_msa_w_centroids %>%
  st_drop_geometry() %>%
  left_join(ok_centroids_geom, by = c("drive_centroid_id" = "ID")) %>%
  select(GEOID, geometry)

```

```{r export to csv for cleaner files, include=FALSE}

write_csv(full_TTmatrix_long, "full_travel_matrix.csv")

st_write(ok_msa, "ok_geom.geojson")

st_write(complete_centroid, "full_centroid_geom.geojson")

```


