---
title: "Trip Generation Model"
author: "Gabriel Barrett-Jackson, Matt Khinda, Samantha Page"
date: "March 2, 2023"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning=FALSE)
```

```{r libraries, echo=FALSE}
library(here)
library(tidyverse)
library(downloader)
library(tidycensus)
library(sf)
library(geojsonio)
```

```{r load NHTS data, echo=FALSE, include=FALSE}
nhts_path <- "../Travel-Forecasting/nhts_data/"

okc_trips <- here(nhts_path,
                  "trippub.csv") %>%
  read_csv() %>%
  filter(HH_CBSA == "36420") %>%
  filter(TRPTRANS != "01" & # Walk
           TRPTRANS != "02" & # Bike
           TRPTRANS != "19") %>%
  mutate(home_based = (WHYFROM == "01" |
                         WHYFROM == "02" |
                         WHYTO == "01" |
                         WHYTO == "02"),
         work = (WHYFROM == "03" |
           WHYTO == "03")) %>%
  mutate(purpose = case_when(home_based & work ~ "HBW",
                             home_based ~ "HBO",
                             TRUE ~ "NHB"))

table(okc_trips$purpose)
```


```{r count household trips, echo=FALSE}
hh_trip_counts <- okc_trips %>%
  group_by(HOUSEID) %>%
  summarise(HBO_trips = sum(purpose == "HBO"),
            HBW_trips = sum(purpose == "HBW"),
            NHB_trips = sum(purpose == "NHB")) 
```

```{r read in nhts data, echo=FALSE}
okc_hhs <- here(nhts_path,
                     "hhpub.csv") %>%
  read_csv() %>%
  filter(HH_CBSA == "36420") %>%
  left_join(hh_trip_counts) %>%
  replace_na(list(HBO_trips = 0,
                  HBW_trips = 0,
                  NHB_trips = 0))
```

```{r variable mutation, echo=FALSE}
okc_trips_65 <- okc_trips %>%
  group_by(HOUSEID) %>%
  summarize(age_65 = sum(ifelse(R_AGE_IMP >= 65, TRUE, FALSE))) %>%
  mutate(age_65 = ifelse(age_65 == 0, FALSE, TRUE))

okc_hhs <- okc_hhs %>%
  mutate(inc_k = case_when(HHFAMINC == "01" ~ 5,
                           HHFAMINC == "02" ~ 12.5,
                           HHFAMINC == "03" ~ 17.5,
                           HHFAMINC == "04" ~ 20,
                           HHFAMINC == "05" ~ 37.5,
                           HHFAMINC == "06" ~ 62.5,
                           HHFAMINC == "07" ~ 82.5,
                           HHFAMINC == "08" ~ 112.5,
                           HHFAMINC == "09" ~ 132.5,
                           HHFAMINC == "10" ~ 175,
                           HHFAMINC == "11" ~ 300,
                           TRUE ~ 99)) %>%
  mutate(inc_k = na_if(inc_k, 99)) %>%
  mutate(have_veh = ifelse(HHVEHCNT == 0, FALSE, TRUE)) %>%
  mutate(homeownership = ifelse(HOMEOWN == "01", TRUE, FALSE))

okc_hhs <- okc_hhs %>%
  left_join(okc_trips_65, by = "HOUSEID") %>%
  mutate(any_kids = LIF_CYC != "01" &
                    LIF_CYC != "02" &
                    LIF_CYC != "09" &
                    LIF_CYC != "10")

```

We ran linear regression models using the following variables: 

* Log-base 2 transformation of median income (in thousands)
* Presence of children in the household
* Homeownership
* Households with residents age 65 or older

```{r Home-Based Other Production Model}
hbo_trip_prod <- lm(HBO_trips ~ log2(inc_k) +
                                #have_veh +
                                any_kids +
                      homeownership +
                      age_65,
                    data = okc_hhs)

summary(hbo_trip_prod)
```
For home-based other trip productions, income, presence of children, and adults over 65 were significant variables in the regression. 



```{r Home-Based Work Production Model}
hbw_trip_prod <- lm(HBW_trips ~ log2(inc_k) +
                                #have_veh +
                                any_kids +
                      homeownership +
                      age_65,
                    data = okc_hhs)

summary(hbw_trip_prod)
```
For home-based work trip productions, only the presence of residents over 65 was significant.

```{r Non-Home Based Production Model}
nhb_trip_prod <- lm(NHB_trips ~ log2(inc_k) +
                                #have_veh +
                                any_kids +
                      homeownership +
                      age_65,
                    data = okc_hhs)

summary(nhb_trip_prod)
```
For non-home-based trip productions, only the presence of children was a significant variable in the regression. 

```{r reading in tract info, include=FALSE, warning=FALSE, message=FALSE}
vars <- c(tot_hhs = "B11005_001",
          hhs_with_kids = "B11005_002",
          hhs_0_veh = "B08201_002",
          hhs_1_veh = "B08201_003",
          med_inc = "B19113_001",
          owner_occ = "B25003_002",
          renter_occ = "B25003_003")

median_study_area_income <- get_acs(year = 2019,
                                    geography = "cbsa",
                                    variables = "B19113_001") %>%
  filter(NAME == "Oklahoma City, OK Metro Area")

# Load TAZs
okc_zones_temp <- get_acs(year = 2019, 
                     state = "OK",
                     county = c("Canadian",
                                    "cleveland",
                                    "Grady",
                                    "Lincoln",
                                    "Logan",
                                    "McClain",
                                    "Oklahoma"),
                     variables = vars,
                     geography = "tract",
                     geometry = TRUE,
                     output = "wide") %>%
  replace_na(replace = list(med_incE = median_study_area_income$estimate[1])) %>%
  select(GEOID, hhs_with_kidsE, med_incE, owner_occE, renter_occE) %>%
  st_drop_geometry()


full_tract_info <- geojson_sf("ok_geom.geojson") %>%
  mutate(pct_veh = (1-(no_vehE/total_hhsE))) %>%
  left_join(okc_zones_temp, by = "GEOID") %>%
  mutate(hhs_without_kidsE = (total_hhsE - hhs_with_kidsE))
```

```{r calculate trip productions, echo=FALSE}
tract_trip_info <- full_tract_info %>%
  mutate(hbo_prod_per_hh = hbo_trip_prod$coefficients["(Intercept)"] + 
                           hbo_trip_prod$coefficients["log2(inc_k)"] * log2(med_incE) + 
                           hbo_trip_prod$coefficients["any_kidsTRUE"] *  (hhs_with_kidsE/total_hhsE) + 
                           hbo_trip_prod$coefficients["homeownershipTRUE"] * (owner_occE / total_hhsE) +
                           hbo_trip_prod$coefficients["age_65TRUE"] * (hh_65plusE / total_hhsE)) %>%
  mutate(hbo_trip_prod = total_hhsE * hbo_prod_per_hh)


tract_trip_info <- tract_trip_info %>%
  mutate(hbw_prod_per_hh = hbw_trip_prod$coefficients["(Intercept)"] + 
                           hbw_trip_prod$coefficients["log2(inc_k)"] * log2(med_incE) + 
                           hbw_trip_prod$coefficients["any_kidsTRUE"] *  (hhs_with_kidsE/total_hhsE) + 
                           hbw_trip_prod$coefficients["homeownershipTRUE"] * (owner_occE / total_hhsE) +
                           hbw_trip_prod$coefficients["age_65TRUE"] * (hh_65plusE / total_hhsE)) %>%
  mutate(hbw_trip_prod = total_hhsE * hbw_prod_per_hh)


tract_trip_info <- tract_trip_info %>%
  mutate(nhb_prod_per_hh = nhb_trip_prod$coefficients["(Intercept)"] + 
                           nhb_trip_prod$coefficients["log2(inc_k)"] * log2(med_incE) + 
                           nhb_trip_prod$coefficients["any_kidsTRUE"] *  (hhs_with_kidsE/total_hhsE) + 
                           nhb_trip_prod$coefficients["homeownershipTRUE"] * (owner_occE / total_hhsE) +
                           nhb_trip_prod$coefficients["age_65TRUE"] * (hh_65plusE / total_hhsE)) %>%
  mutate(nhb_trip_prod = total_hhsE * nhb_prod_per_hh)
```

```{r trip attractions, echo=FALSE}
tract_trip_info <- tract_trip_info %>%
  mutate(hbo_trip_attr = 1.0 * total_hhsE +
                         0.3 * basic_emp +
                         5.9 * retail_emp +
                         2.3 * service_emp) %>%
  mutate(hbw_trip_attr = 1.2 * total_emp) %>%
  mutate(nhb_trip_attr = 0.6 * total_hhsE +
                         0.7 * basic_emp +
                         2.6 * retail_emp +
                         1.0 * service_emp)
```

```{r comparing productions and attractions, include=FALSE}
sum(tract_trip_info$hbo_trip_prod)

sum(tract_trip_info$hbo_trip_attr)

tract_trip_info <- tract_trip_info %>%
  mutate(hbo_bal_attr = hbo_trip_attr * 
           sum(hbo_trip_prod) / sum(hbo_trip_attr))

sum(tract_trip_info$hbo_bal_attr)


sum(tract_trip_info$hbw_trip_prod)

sum(tract_trip_info$hbw_trip_attr)

tract_trip_info <- tract_trip_info %>%
  mutate(hbw_bal_attr = hbw_trip_attr * 
           sum(hbw_trip_prod) / sum(hbw_trip_attr))

sum(tract_trip_info$hbw_bal_attr)



sum(tract_trip_info$nhb_trip_prod)

sum(tract_trip_info$nhb_trip_attr)

tract_trip_info <- tract_trip_info %>%
  mutate(nhb_bal_attr = nhb_trip_attr * 
           sum(nhb_trip_prod) / sum(nhb_trip_attr))

sum(tract_trip_info$nhb_bal_attr)
```
```{r write a new file, echo=FALSE, warning=FALSE}
#tract_trip_info %>%
  #st_write("tract_trip_info.geojson")
```

As expected, attractions and productions are primarily concentrated in the downtown area--attractions even more so than productions. Both generally follow the density maps of residences and employment that we created earlier in the semester. 

```{r dot density map on productions, echo=FALSE}
#productions dot density map
hbo_prod_pts <- st_sample(tract_trip_info, 
                              size = ceiling(tract_trip_info$hbo_trip_prod/1000))
 
hbo_prod_df <- tibble(trip_type = rep("Home-Based Other",
                                            length(hbo_prod_pts))) %>%
  st_sf(geom = hbo_prod_pts)

hbw_prod_pts <- st_sample(tract_trip_info,
                               size = ceiling(tract_trip_info$hbw_trip_prod/1000))

hbw_prod_df <- tibble(trip_type = rep("Home-Based Work", 
                                              length(hbw_prod_pts))) %>%
  st_sf(geom = hbw_prod_pts)

nhb_prod_pts <- st_sample(tract_trip_info,
                               size = ceiling(tract_trip_info$nhb_trip_prod/1000))

nhb_prod_df <- tibble(trip_type = rep("Non-Home-Based", 
                                              length(nhb_prod_pts))) %>%
  st_sf(geom = nhb_prod_pts)

production_points <- rbind(hbo_prod_df, hbw_prod_df, nhb_prod_df)


ggplot(full_tract_info) +
  geom_sf(color = "white") +
  geom_sf(data = production_points,
          aes(color = trip_type),
          alpha = 0.5,
          size = 0.4) +
  scale_color_brewer("Production Distribution\n(each points represents\n1000 productions)",
                     palette = "Set2") +
  theme_void()  +
  guides(color = guide_legend(override.aes = list(size=5, alpha = 0.6)))
```

```{r dot density map for attractions, echo=FALSE}
#attractions
hbo_att_pts <- st_sample(tract_trip_info, 
                              size = ceiling(tract_trip_info$hbo_trip_attr/1000))
 
hbo_att_df <- tibble(trip_type = rep("Home-Based Other",
                                            length(hbo_att_pts))) %>%
  st_sf(geom = hbo_att_pts)

hbw_att_pts <- st_sample(tract_trip_info,
                               size = ceiling(tract_trip_info$hbw_trip_attr/1000))

hbw_att_df <- tibble(trip_type = rep("Home-Based Work", 
                                              length(hbw_att_pts))) %>%
  st_sf(geom = hbw_att_pts)

nhb_att_pts <- st_sample(tract_trip_info,
                               size = ceiling(tract_trip_info$nhb_trip_attr/1000))

nhb_att_df <- tibble(trip_type = rep("Non-Home-Based", 
                                              length(nhb_att_pts))) %>%
  st_sf(geom = nhb_att_pts)

attraction_points <- rbind(hbo_att_df, hbw_att_df, nhb_att_df)


ggplot(full_tract_info) +
  geom_sf(color = "white") +
  geom_sf(data = attraction_points,
          aes(color = trip_type),
          alpha = 0.5,
          size = 0.4) +
  scale_color_brewer("Attraction Distribution\n(each points represents\n1000 attractions)",
                     palette = "Set1") +
  theme_void()  +
  guides(color = guide_legend(override.aes = list(size=5, alpha = 0.6)))
```